---
title: "Working with tidy data: Mutate"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(RCurl)
library(lubridate)
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE)
# gradethis::gradethis_setup()

fractured_fruit <-
  tibble(
    fruit_prefix = fruit[1:3] %>%
      str_sub(1,3), 
    fruit_suffix = fruit[1:3] %>%
      str_sub(start =  4),
    fruit_type = c('pome', 'drupe', 'berry'),
    consumption_class = c('clean', 'sloppy', 'sloppy'))

long_trees <-
  tibble(
    tree_id = 1:5,
    species = c('Carya glabra', 'Quercus rubra', 'Carya glabra', 'Acer saccharum', 'Acer saccharum'),
    dbh = c(8, 32, 27, 24, 42),
    height = c(9, 23, 24, 18, 26)) %>% 
  pivot_longer(dbh:height, names_to = 'measure', values_to = 'value')

addresses <-
  tibble(
    name_last = c('Bonacci', 'Wickham', 'MacArthur', 'Euler'),
    name_first = c('Leonardo', 'Hadley', 'Robert', 'Leonhard'),
    address = 
      c('2358 Rabbit Lane ,Pisa,Italy',
        '2014 Tidy Cat Boulevard,Aukland, New Zealand',
        '1958 Warbler Court, Princeton,New Jersey',
        '7 Bridges Path,Konigsberg,Russia')) %>% 
  arrange(name_last)

my_covid_garden_sites <-
  tibble(
    bed = 1:5,
    light_access = c('full sun', 'full sun' , 'partial shade', 'full sun', 'shade'),
    plants = c('jalepenos', 'tomatoes', 'kale', 'summer squash', 'mesclun'))

my_covid_garden_events <-
  tribble(
    ~bed, ~date, ~event,
    1, '2020-04-12', 'planted',
    1, '2020-05-17', 'weeded',
    1, '2020-07-16', 'fenced',
    1, '2020-09-23', 'harvested',
    2, '2020-04-14', 'planted',
    2, '2020-05-12', 'weeded',
    2, '2020-07-14', 'eaten by deer',
    3, '2020-04-16', 'planted',
    3, '2020-05-26', 'weeded',
    3, '2020-06-23', 'eaten by slugs',
    3, '2020-07-12', 'applied diatomaceous earth',
    3, '2020-08-07', 'harvested',
    4, '2020-04-17', 'planted',
    4, '2020-05-12', 'weeded',
    4, '2020-07-14', 'trampled by deer',
    5, '2020-04-03', 'planted',
    5, '2020-05-19', 'weeded',
    5, '2020-05-20', 'eaten by squirrels') %>% 
  arrange(date) %>% 
  mutate(date = as_date(date))

my_covid_garden <-
  left_join(
    my_covid_garden_sites,
    my_covid_garden_events,
    by = 'bed')

rm(my_covid_garden_sites)

rm(my_covid_garden_events)


urban_sites <-
  read_csv('../data/sites.csv')

messy_weather <-
  read_csv('../data/messy_weather.csv')

messy_weather_long_days <-
  pivot_longer(
    data = messy_weather,
    cols = march_1:march_31,
    names_to = "day",
    values_to = "value",
    names_prefix = "march_")

messy_weather_date_fix <-
  unite(
    messy_weather_long_days,
    col = 'date',
    c('year', 'month', 'day'),
    sep = '-')

messy_weather_wide_weather <-
  messy_weather_date_fix %>% 
  pivot_wider(
    names_from = variable,
    values_from = value)

messy_weather_temperature_fix <-
  separate(
    messy_weather_wide_weather,
    col = temperature_min_max,
    into = c('temperature_min', 'temperature_max'),
    sep = ':')

weather_stations <-
  distinct(
    select(
      messy_weather_temperature_fix,
      station:name))

observations <-
  select(
    messy_weather_temperature_fix,
    station, date:temperature_max)

monarch_inat <-
  read_csv('../data/monarch_inat.csv')

multiply_by_two <- 
  function(x) {
    x*2
  }

add_one <- 
  function(x) {
    x+1
  }

messy_weather_temperature_fix <-
  messy_weather %>% 
  pivot_longer(
    cols = march_1:march_31, 
    names_to = 'day', 
    values_to = 'value', 
    names_prefix = 'march_') %>% 
  unite(
    c('year', 'month', 'day'), 
    col = 'date', 
    sep = '-',
    na.rm = TRUE) %>% 
  pivot_wider(
    names_from = variable, 
    values_from = value) %>% 
  separate(
    col = temperature_min_max,
    into = c('temperature_min', 'temperature_max'),
    sep = ':')


weather_stations <-
  messy_weather_temperature_fix %>% 
  select(station:name) %>% 
  distinct()

observations <-
  messy_weather_temperature_fix %>% 
  select(station, date:temperature_max)
```


<head>
<link rel="stylesheet" href="https://cdnjs.cloudflare.com/ajax/libs/font-awesome/4.7.0/css/font-awesome.min.css">
<link rel="stylesheet" type="text/css" href="../css/styles.css">
<!-- link rel="stylesheet" href="https://maxcdn.bootstrapcdn.com/bootstrap/3.3.7/css/bootstrap.min.css" -->
<style>

body {
  font-size: 18px;
}

.topic {
  font-size: 19px !important;
}

code {
  font-size: 16px !important;
}

h1 {
  font-size: 2.25em;
}

h2 {
  font-size: 2em;
}

h3 {
  font-size: 1.75em;
}

h4 {
  font-size: 1.5em;
}

h5 {
  font-size: 1.25em;
}

pre {
  font-size: 16px !important;
  margin-top: 2em;
  margin-bottom: 2em;
}
</style>
</head>

## Overview

<a title = "This is not a pipe">
  <img  src = "https://upload.wikimedia.org/wikipedia/en/b/b9/MagrittePipe.jpg"
        alt="Magritte painting, the treachery of images"
        style = "float: right; padding-left: 3em; padding-bottom: 3em; width: 50%; height: auto;">
  </img>
</a>

- **Review**: Object assignment and mutation in base R.
- **Mutation fundamentals**: Develop a fundamental understanding of how, when, and why to mutate.
- **Mutating with logic**: Some special, oft-used, applications of functions inside of mutate.</li>

<hr>

## Review: Modifying data frames

Consider the line of code below:

```{r, echo = TRUE}
df <-
  tibble(
    j = 'hello', 
    h = 'world')
```

<div class = "reviewDiv" style = "margin-top: 2em;">
<h4><i class="fa fa-user-circle" aria-hidden="true" style = "font-size: 100%"></i> Creating and assigning objects</h4> 
```{r question_ymd, echo=FALSE}
question("Which of the following is true about this section of code (choose one or more)?",
  answer("The assignment operature = is a primitive function", correct = TRUE, message = "There are four functions in this section of code. The assignment operators <- and = are primitive functions and tibble is an R function. Objects j and h are vectors stored inside df but are not directly accessible from the parent environment."),
  answer("The code contains one function"),
  answer("The assignment operator <- is a primitive function", correct = TRUE),
  answer("Object j is a vector stored in the same environment that contains df"),
  type = 'multiple',
  allow_retry = TRUE,
  random_answer_order = TRUE
)
```
</div>

If we wanted to modify a column in the data frame using base R, we use the primitive function `$` to reference the column and the primitive function `<-` to assign a vector of new values:

```{r, echo = TRUE}
df <-
  tibble(
    j = 'hello', 
    h = 'world')

df$h <- 'howdy'

df
```

The vector `h` exists inside the environment of the data frame object `df`. It is not directly accessible from the global environment:

```{r, echo = TRUE}
df <-
  tibble(
    j = 'hello', 
    h = 'world')

df$h <- 'howdy'

exists('h')

df
```

This is why the `$` operator is used. It references a child object, by name, within the environment of its parent.

We can also use this to add a new child object to the parent. Let's append `df`, adding the column `i`:

```{r, echo = TRUE}
df <-
  tibble(
    j = 'hello', 
    h = 'world')

df$i <- 'howdy'

df
```

You can also use the `$` operator and the assignment operator to remove a columm from the data frame:


```{r, echo = TRUE}
df <-
  tibble(
    j = 'hello', 
    h = 'world',
    i = 'howdy')

df$i <- NULL

df
```

<hr>

## Introduction to mutate

The `mutate` family of functions in tidyverse (dplyr) gives you the power to add, modify, and delete variables within a data frame. 

The first argument of `mutate` is the data frame you would like to modify. Let's create a data frame as above and add a column using mutate:

```{r, echo = TRUE}
df <-
  tibble(
    j = 'hello', 
    h = 'world')

mutate(df, i = 'howdy')
```

In a nested function structure, this would look like this:

```{r, echo = TRUE}
mutate(
  tibble(
    j = 'hello', 
    h = 'world'),
  i = 'howdy')
```

Of course, using the pipe operator `%>%` (CMD+Shift+M or CTRL+Shift+M) would be a more modern way to go about this. Repeat the above using a pipe (do not assign anything):

```{r mutate_start, exercise = TRUE}

```

```{r mutate_start-hint-1}
tibble(
  j = 'hello', 
  h = 'world')
```

```{r mutate_start-hint-2}
tibble(
  j = 'hello', 
  h = 'world') %>% 
  mutate(i = 'howdy')
```

We can also use `mutate` to modify an existing column:

```{r, echo = TRUE}
mutate(
  tibble(
    j = 'hello', 
    h = 'world'),
  h = 'howdy')
```

A benefit to `mutate` is that you choose when you want to modify a variable in your global environment. While this is great from a modern coding perspective, if you're used to assigning intermediate objects, can lead to a lot of head scratching. Note the behavior below:

```{r, echo = TRUE}
df <-
  tibble(
    j = 'hello', 
    h = 'world')

mutate(
  df,
  j = 'boy',
  h = 'howdy')

df
```


### The environment of mutate

Mutate works within the environment of the data frame being evaluated. This saves lots of writing.

Consider the data frame below. Use base R to add one to each value:

Use base R to create a new column, "c", that multiplies column `b` by two:

```{r no_dollar_base, exercise = TRUE}
example <-
  tribble(
    ~ a, ~ b,
    1, 1,
    2, 3)
```

```{r no_dollar_base-hint-1}
example <-
  tribble(
    ~ a, ~ b,
    1, 1,
    2, 3)

example$c
```

```{r no_dollar_base-hint-2}
example <-
  tribble(
    ~ a, ~ b,
    1, 1,
    2, 3)

example$c <- example$b
```

```{r no_dollar_base-hint-3}
example <-
  tribble(
    ~ a, ~ b,
    1, 1,
    2, 3)

example$c <- example$b * 2
```

Because `mutate` already works within the environment of the data frame we can remove the `$`:

```{r, echo = TRUE}
tribble(
    ~ a, ~ b,
    1, 1,
    2, 3) %>% 
  mutate(c = b * 2)
```

Consider this data frame of iNaturalist observations of Monarch butterflies. The data contain the summed annual observations for a given year and city:

```{r, echo = TRUE}
monarch_inat
```

If we wanted to calculate the total across years for each city. In base R this would look like this:

```{r, echo = TRUE}
monarch_inat$total <-
  monarch_inat$year_2018 +
  monarch_inat$year_2019 +
  monarch_inat$year_2020

monarch_inat
```

Using a pipe and mutate, create a new column called "total" that is the sum across years for each city:

```{r monarch_inat_sum2, exercise = TRUE}

```

```{r monarch_inat_sum2-hint-1}
monarch_inat %>% 
  mutate()
```

```{r monarch_inat_sum2-hint-2}
monarch_inat %>% 
  mutate(total = )
```

```{r monarch_inat_sum2-hint-3}
monarch_inat %>% 
  mutate(total = year_2018 + year_2019 + year_2020)
```

### Grouping inputs by rows

Why doesn't `sum` work? Let's look at the results of sum:

```{r, echo = TRUE}
monarch_inat %>% 
  mutate(total = sum(year_2018, year_2019, year_2020))
```

Luckily, dplyr has a function called `rowwise` that evaluates a function along a single row at a time. We could apply it here using:

```{r, echo = TRUE}
monarch_inat %>% 
  rowwise() %>% 
  mutate(total = sum(year_2018, year_2019, year_2020))
```

The function `rowwise` works by breaking a data frame into a group and applying a function to the grouped data. Whenever you do a grouped calculate it's critical to use the function `ungroup` when the grouping is no longer needed:

```{r, echo = TRUE}
monarch_inat %>% 
  rowwise() %>% 
  mutate(total = sum(year_2018, year_2019, year_2020)) %>% 
  ungroup()
```

<hr>

## Multiple mutations

What if we wanted to add or modify multiple variables? In base R, this would look like:

```{r, echo = TRUE}
df <-
  tibble(
    j = 'hello', 
    h = 'world')

df$j <- 'boy'

df$h <- 'howdy'

df
```

Using `mutate` we simply separate our mutated variables with a comma:

```{r, echo = TRUE}
tibble(
  j = 'hello', 
  h = 'world') %>% 
  mutate(
    j = 'boy',
    h = 'howdy')
```

### Sequential mutation

Operations can be chained together such that subsequent mutate statements are dependent on the previous value.

Using our monarch iNaturalist data, if we wanted to calculate the proportional contribution of a given year's counts to the total, we could calculate the total across years, as you did above, and pipe the resultant value into a new mutate statement. Here's how it would look for 2020:


```{r, echo = TRUE}
monarch_inat %>% 
  rowwise() %>% 
  mutate(total = sum(year_2018, year_2019, year_2020)) %>% 
  ungroup() %>% 
  mutate(year_2020 = year_2020/total)
```

We could also place both operations in the same mutate statement:

```{r, echo = TRUE}
monarch_inat %>% 
  rowwise() %>% 
  mutate(
    total = sum(year_2018, year_2019, year_2020),
    year_2020 = year_2020/total) %>% 
  ungroup()
```

### Applying the same function to multiple variables

What if we wanted to mutate multiple variables in the same way? If we wanted to repeat the above to see the proportional contribution of year, we could write it like this:

```{r, echo = TRUE}
monarch_inat %>% 
  rowwise() %>% 
  mutate(
    total = sum(year_2018, year_2019, year_2020),
    year_2018 = year_2018/total,
    year_2019 = year_2019/total,
    year_2020 = year_2020/total) %>% 
  ungroup()
```

That's a lot of repetition though, and repetitive code is dangerous code.

We can use `mutate_at` to avoid this. The arguments of `mutate_at` include:

* The name of the data frame
* The variables you would like to mutate, specified with `vars(name1, name2)`
* a right hand formula specified with `~` that uses the wildcard `.` in place of the variable name

Here's how it would look:

```{r, echo = TRUE}
monarch_inat %>% 
  rowwise() %>% 
  mutate(total = sum(year_2018, year_2019, year_2020)) %>% 
  ungroup() %>% 
  mutate_at(
    vars(year_2018:year_2020),
    ~ ./total)
```

### Conditional mutation

Perhaps you are interested in mutating every numeric variable in the same way. To address this, you can use the function `mutate_if`. The arguments of `mutate_if` include:

* The name of the data frame
* A logical test for the variable to include (`TRUE` are evaluated)
* a right hand formula specified with `~` that uses the wildcard `.` in place of the variable name

Let's calculate the proportional contribution of each city to the total annual observations of monarchs:

```{r, echo = TRUE, eval = TRUE}
monarch_inat %>% 
  mutate_if(is.numeric, ~ ./sum(.))
```

Calculate the proportional contribution of each city to monarch for each year and across years:

```{r monarch_final, exercise = TRUE}

```

```{r monarch_final-hint-1}
monarch_inat
```

```{r monarch_final-hint-2}
monarch_inat %>%
  rowwise()
```

```{r monarch_final-hint-3}
monarch_inat %>%
  rowwise() %>% 
  mutate(total = sum(year_2018, year_2019, year_2020))
```

```{r monarch_final-hint-4}
monarch_inat %>%
  rowwise() %>% 
  mutate(total = sum(year_2018, year_2019, year_2020)) %>% 
  ungroup()
```

```{r monarch_final-hint-5}
monarch_inat %>%
  rowwise() %>% 
  mutate(total = sum(year_2018, year_2019, year_2020)) %>% 
  ungroup() %>%  
  mutate_if(is.numeric, ~ ./sum(.))
```

## Transmutation

The function `transmute` is very similar to mutate, but with one important distinction - only the columns specified in the transmute statement are maintained.

Let's use the last exercise as an example:

```{r, echo = FALSE, eval = TRUE}
monarch_inat %>%
  rowwise() %>% 
  mutate(total = sum(year_2018, year_2019, year_2020)) %>% 
  ungroup() %>%  
  mutate_if(is.numeric, ~ ./sum(.))
```

If we are only interested in looking at the total by city, there's more output than we need. Even removing the final `mutate_if` statement yield more data than we're interested in:

```{r, echo = FALSE, eval = TRUE}
monarch_inat %>%
  rowwise() %>% 
  mutate(total = sum(year_2018, year_2019, year_2020)) %>% 
  ungroup()
```

If we use the function `transmute` rather than `mutate` we can specify out columns of interest:

```{r, echo = FALSE, eval = TRUE}
monarch_inat %>%
  rowwise() %>% 
  transmute(
    city,
    total = sum(year_2018, year_2019, year_2020)) %>% 
  ungroup()
```

This makes `transmute` very powerful if the columns themselves are intermediate data!

<hr>

## Mutate with logic

Two functions that I use inside of mutate often are `if_else` and `case_when` and are worth giving special attention here. Both perform logical tests on data and reclassify data based on that test.

### if_else

The tidyverse function `if_else` takes as its arguments:

* A logical test 
* Value if TRUE
* Value if FALSE
* Value if missing (optional, defaults to NULL)

```{r, eval = TRUE, echo = TRUE}
color <- 
  c('vermilion', 'cobalt')

if_else(
  color == 'cobalt',
  'blue',
  'red')
```

The tidyverse function `if_else` is slightly different than the base R function `ifelse`. It requires that the output is of the same data type:

```{r, eval = TRUE, echo = TRUE, error = TRUE}
color <- 
  c('vermilion', 'cobalt')

if_else(
  color == 'cobalt',
  NA,
  'red')
```

```{r, eval = TRUE, echo = TRUE, error = TRUE}
color <- 
  c('vermilion', 'cobalt')

ifelse(
  color == 'cobalt',
  NA,
  'red')
```

To address this, you have to specify how the NA is encoded (e.g., NA_integeter_, NA_real_):

```{r, eval = TRUE, echo = TRUE, error = TRUE}
color <- 
  c('vermilion', 'cobalt')

if_else(
  color == 'cobalt',
  NA_character_,
  'red')
```

Sometimes you only want to change one of the values:

```{r, eval = TRUE, echo = TRUE}
color <- 
  c('vermilion', 'cobalt')

if_else(
  color == 'cobalt',
  'blue',
  color)
```

Here's how you would use an `if_else` statement inside of mutate:

```{r, eval = TRUE, echo = TRUE}
tibble(
  color = c('vermilion', 'cobalt')) %>% 
  mutate(
    color = if_else(
      color == 'cobalt',
      'blue',
      'red'))
```

It's sometimes useful to use evaluate a function inside of the logical test of your `if_else` statement. For example, you can classify a set of values as "high" or "low" using:

```{r, eval = TRUE, echo = TRUE}
tibble(
  value = 1:10) %>% 
  mutate(
    value_class = if_else(
      value >= mean(value),
      'high',
      'low'))
```

### case_when

If you have to conduct multiple tests, the tidyverse function `case_when` is your best option. The syntax for `case_when` is a series of two-sided formulas, where the left-hand side is the logical test and the right is the replacement values.

```{r, eval = TRUE, echo = TRUE}
x <- 1:10 

case_when(
  x < 4 ~ 'small',
  x > 7 ~ 'large')
```

The left-hand side of the formula evaluates to TRUE or FALSE and the right-hand side tells R what to do when this is the case.
Notice that the value "5" became an `NA` because it did not fit either condition. We can specify this middle condition as follows:

```{r, eval = TRUE, echo = TRUE}
x <- 1:10 

case_when(
  x < 4 ~ 'small',
  x > 7 ~ 'large',
  TRUE ~ 'medium')
```

This might be safe, but it's better to specify the values to map:

```{r, eval = TRUE, echo = TRUE}
x <- 1:10 

case_when(
  x <= 3 ~ 'small',
  x <= 7 ~ 'medium',
  x > 7 ~ 'large')
```

Here's how you would use `case_when` inside of mutate:

```{r, eval = TRUE, echo = TRUE}
tibble(x = 1:10) %>% 
  mutate(
    x = case_when(
      x <= 3 ~ 'small',
      x <= 7 ~ 'medium',
      x > 7 ~ 'large'))
```

Similar to the `if_else` statement, it can also be useful to use a function to classify data in a `case_when` statement. The function `stats:::quantile` can be used to calculate sample quantiles:

```{r, eval = TRUE, echo = TRUE}
tibble(x = 1:10) %>% 
  mutate(
    x = case_when(
      x <= quantile(x, .25) ~ 'small',
      x < quantile(x, .75) ~ 'medium',
      x >= quantile(x, .75) ~ 'loud'))
```

**Now you**

The data frame `urban_sites` (Pharr et al. *in prep*), provides information on the proportion of impervious surface, noise pollution, and light pollution at a series of sites within metropolitan Washington, DC. 

```{r, eval = TRUE, echo = TRUE}
glimpse(urban_sites)
```


Modify the data set such that: 

* Include a column called "urban_class". Classify sites as urban if they are greater than 60% impervious surface, rural if they are below 10%, and suburban anywhere in between.
* Modify the column `noise_pollution`. If values are greater than the mean, call it "noisy" and less-than-or-equal to the mean, call the site "quiet".
* Modify the column `light_pollution`. If values are in the upper quartile, call the values "light", if in the lower quartile, call the values "dark", call all other values the values "medium".

```{r urban_light_noise, exercise = TRUE}

```

```{r urban_light_noise-hint-1}
urban_sites %>% 
  mutate(
    urban_class = ,
    noise_pollution = ,
    light_pollution = )
```

```{r urban_light_noise-hint-2}
urban_sites %>% 
  mutate(
    urban_class = 
      case_when(),
    noise_pollution = 
      if_else(),
    light_pollution = 
      case_when())
```

```{r urban_light_noise-hint-3}
urban_sites %>% 
  mutate(
    urban_class = 
      case_when(
        impervious_surface > 60 ~ 'urban',
        impervious_surface >= 10 ~ 'suburban',
        impervious_surface < 10 ~ 'rural'),
    noise_pollution = 
      if_else(),
    light_pollution = 
      case_when())
```

```{r urban_light_noise-hint-4}
urban_sites %>% 
  mutate(
    urban_class = 
      case_when(
        impervious_surface > 60 ~ 'urban',
        impervious_surface >= 10 ~ 'suburban',
        impervious_surface < 10 ~ 'rural'),
    noise_pollution = 
      if_else(
        noise_pollution > mean(noise_pollution),
        'noisy',
        'quiet'),
    light_pollution = 
      case_when())
```

```{r urban_light_noise-hint-5}
urban_sites %>% 
  mutate(
    urban_class = 
      case_when(
        impervious_surface > 60 ~ 'urban',
        impervious_surface >= 10 ~ 'suburban',
        impervious_surface < 10 ~ 'rural'),
    noise_pollution = 
      if_else(
        noise_pollution > mean(noise_pollution),
        'noisy',
        'quiet'),
    light_pollution = 
      case_when(
        light_pollution >= quantile(light_pollution, .75) ~ 'light',
        light_pollution <= quantile(light_pollution, .25) ~ 'quiet',
        light_pollution > quantile(light_pollution, .25) ~ 'medium'))
```

**fin!**

<hr>
